Detecting the impact of land cover change on observed rainfall.

### 2019-02-28 Manuscripts/PeerJRainfall_Landuse/Rainfall_Landuse.tex
update for PeerJ paper

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Journal of Hydrology Regional Studies} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Detecting influences of changes in land cover on observed rainfall.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)
% Pandoc header

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\begin{frontmatter}

  \title{Detecting influences of changes in land cover on observed rainfall.}
    \author[a]{Chun X. Liang}
   \ead{summer.chunliang@gmail.com}

    \author[a]{Floris F. van Ogtrop}
   \ead{floris.vanogtrop@sydney.edu.au}

    \author[a]{R. Willem Vervoort\corref{c1}}
   \ead{willem.vervoort@sydney.edu.au} 
   \cortext[c1]{Corresponding Author}
      \address[a]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}

  \begin{abstract}
  Analysis of observational data to identify relationships between
  rainfall and land cover change are difficult due to multiple
  environmental factors that cannot be strictly controlled. In this study
  we present a methodology to investigate the relationship using
  statistical methods on data from best available sources at two sites in
  Australia. Gridded data of rainfall and tree cover were used as
  spatially corresponding local conditions. Large scale effects were
  represented by climatic indicators, such as SOI and IOD. Regression
  analysis and step trend tests were used to assess the effect of abrupt
  land surface intervention. At a Queensland site, significant tree cover
  change between 2002 - 2005 did not result in strong statistically
  significant precipitation changes. On the other hand, results from a
  bushfire affected NSW/VIC region suggests significant changes in the
  rainfall. This indicates the method works better when a abrupt change in
  the data can be clearly identified. The results from the step trend test
  implied a positive relationship between the tree cover and the rainfall
  at 0.1 significance level in both locations in data up to 2009. However,
  high rainfall variability and possible regrowth meant that no signifcant
  changes were observed in alonger time sereies to 2015.
  \end{abstract}

 \end{frontmatter}

\section{Introduction}\label{introduction}

Land use and land cover changes can lead to changes in the local
climate. Empirical and modelling studies have found cloud types and
rainfall are correlated to large scale vegetation cover changes, such as
deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
measurement in Western Australia, Junkermann et al. (2009) showed a
significantly higher level of aerosols over an agricultural area
compared to an adjacent natural vegetation. They suggested that a
modification of aerosol concentrations due to deforestation could have
contributed to a reduction of local rainfall, as more, but smaller rain
droplets were observed. Nair et al. (2011) reported from the Bunny Fence
Experiment in Western Australia that local land use change altered the
synoptic west coast trough dynamics and surface roughness, and this
resulted in an observed rainfall decrease. Maximum temperatures were
also found to be sensitive to land cover change in eastern Australia
(McAlpine et al. 2007).

Overall the number of empirical studies analyzing changes to rainfall
due to land cover change from observational data is limited. Most of the
studies mentioned previously were either model simulations, or
comparisons of modelled data with observations. This is because there
are some fundamental experimental difficulties in both space (where does
evaporated water reappear as rainfall?) and in time (how much time does
it take for land cover change effects to appear or disappear?). In
addition, in many areas across the globe rainfall variability is related
to complex set of interactions, of which land use change might only be a
minor component.

Locally, there are two main sources that generate rainfall: moisture
from advective atmospheric transport; and local evapotranspiration
(Eltahir and Bras 1996; Bosilovich and Chern 2006; Dirmeyer, Brubaker,
and DelSole 2009; Gimeno et al. 2010). The local evapotranspiration
component is the component considered to be affected by land use change
(Eltahir and Bras 1996). According to Trenberth (1999), the contribution
of advective moisture partially depends on the availability of external
moisture and atmospheric transport. On the longer time scale, such as
monthly and annually, large scale atmospheric dynamics are affected by
large scale climate drivers. For example, many studies have reported
significant relationships between rainfall in large parts of Australia
and the El Niño-Southern Oscillation (ENSO) (Verdon et al. 2004; Risbey
et al. 2009; Speer, Leslie, and Fierro 2011). In contrast, local ET is
determined by local land surface characteristics, which influence local
scale atmospheric dynamics and hence the amount of rainfall, including
contribution from both main sources. Therefore land surface is suggested
to play an important role in local rainfall.

Although climate drivers demonstrate some capability to predict
Australian rainfall, there is still a large amount of unexplained
variance. Westra and Sharma (2010) pointed out that models based on
global sea surface temperature anomalies can only predict up to 14.7\%
of annual precipitation variance. Some of the remaining variance could
be due to land surface processes as suggested in studies predicting
local rainfall (e.g. Ma et al. 2011; Zeng et al. 2012; Pitman and Lorenz
2016; Saha, Dirmeyer, and Chase 2016). However, they are mostly based on
modelling experiments and little evidence was reported from
observations. However, Pitman et al. (2004) found a good match between
observations and simulated rainfall changes in southwest Western
Australia ,forced by land cover change. Timbal and Arblaster (2006) were
able to reproduce the rainfall decline in south west Australia by
including land cover influence. In addition, local land use change might
not be a primary, but is likely to be a secondary cause of rainfall
change (Nicholls 2006).

Therefore, the aim of this study is to expand the available methodology
to use empirical evidence at regional scales to investigate the cause
and effect relationship between land cover change and local rainfall.
More specifically, we hypothesize that a step change on the land cover
on the surface will cause a step change in the rainfall. To demonstrate
the approach and this effect we applied the methodology to study the
changes in rainfall at two locations in Queensland and NSW/Victoria
where there are possible step changes in land cover change due to land
clearing and bush fires. The methodology is mainly based on statistical
approaches to identify changes in rainfall, which were subsequently
associated with land cover change through spatial comparison.

@@

We use two areas in Australia to demonstrate the approach. In Australia,
significant tree cover change has mainly occurred in the north east of
the continent and on the southeast coast, as well as in the southwest of
Western Australia. According to the National Dynamic Land Cover Dataset
(DLCD) (Lymburner et al. 2010), most of these areas have experienced
decreasing EVI post 2000. As an index for vegetation greenness, the
decreasing values indicate lower biomass over time in the tree cover
regions. The possible EVI reduction might be due to land clearing, bush
fires or drought.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/map_selreg} \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest in which tree cover is denser.}\label{fig:selreg}
\end{figure}

Two regions were selected where significant tree cover change was
reported. The first region is located in south central Queensland to the
north of the Murray Darling Basin (MDB) (site 1 in Figure
\ref{fig:selreg}. High rates of land clearing have been reported in this
region during the early 2000s (Department of Natural Resources and Water
2007). The second study region is located at the border of New South
Wales and Victoria, and includes the Snowy Mountain ranges (site 2 in
Figure \ref{fig:selreg}. Severe bush fires occurred in this area and the
surroundings in early 2003 (see Figure \ref{fig:figure2bushfire}). The
2003 bush fires were the largest and the worst in this area for the last
60 years (The State Government of Victoria 2011). Two thirds of
Kosciuszko national park was heavily burned and regrowth was reported to
be slow due to drought and cold conditions (ABC News 2003) and the type
of species in this region. However, in the longer term, after an early
high transpiration period a recovery of pre-fire evapotranspiration
would be expected (Kuczera 1987). As a result , significant tree cover
loss has happened in both study areas in the last decade, either
permanently or temporarily.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows large area in the Kosciuszko national park has been burned. Some locations in the southwest of ACT have also experienced intensive bushfires.}\label{fig:figure2bushfire}
\end{figure}

The two regions have different climate characteristics. The QLD region
is partially grassland and partially subtropical, while the NSW/VIC
region is mainly within the temperate zone, under the Köppen
classification. According to Australian Bureau of Meteorology (BoM), the
NSW/VIC region receives 1000 - 2000 mm rainfall annually, which is more
than double of the rainfall in the QLD region. Evapotranspiration is
similar in both regions. Marine moisture and orographic effects are
likely to be the main contributors to rainfall in the southeast mountain
areas of the NSW/VIC region.

The land use and land cover characteristics in the two regions are also
different. In the Queensland region, the tree cover is sparse over most
of the area. The MODIS satellite tree cover data (discussed in more
detail in section 3) shows that tree cover in this region is generally
below 20\% of total ground area. Grazing is the main activity in this
region, with over 90\% of land used by the grazing industry (ABARES
2010). Our starting assumption is that the main cause of the EVI decline
over large part of the region is due to land clearing. Tree cover has
been cleared at a massive scale over the last decade, especially during
2002 - 2004. The Kosciuszko national park is within the NSW/VIC region.
Here tree cover is denser with open or even closed forest (the tree
cover distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
species in the alpine area are Snow Gum and large stand species such as
Alpine Ash and Mountain Gum in the sub-alpine area. These trees can
reach a great height but they take long time to grow. For example,
Alpine Ash would need about 20 years to mature. Although land clearing
is not a major issue in this region, it is vulnerable to fires and
drought.

Therefore two types of land cover changes were studied. The reports from
the Queensland Statewide Land Cover and Trees Study (SLATS) (e.g.
Department of Natural Resources and Mines 2005; Department of Science,
Information Technology and Innovation 2017) were used to investigate the
time and location of land clearing in the QLD region. The MODIS burned
area product, MCD45A1 (Roy, Lewis, and Justice 2002; Roy et al. 2005;
Roy et al. 2008), was used to locate bush fires areas in the NSW/VIC
region, with a grid resolution of 500 m. MCD45A1 provides monthly
burning information on all pixels, which helps to pinpoint an abrupt
event. Due to the nature of the different land cover change, the
post-change vegetation status in the two regions is expected to be
different (see Figure \ref{fig:figure3-tc-simple}).

@@

The hypothesis in this study therefore is that the effect of 2003 - 2004
land clearings in the QLD region and the 2003 bush fires in the NSW/VIC
region cause a step change in the local rainfall. The actual tree cover
change at the pixel level during this time was derived from the 15-year
MODIS data (discussed below). The difference of tree cover before and
after the land disturbance was tested using a Student's t-test. As the
length of the tree cover data is shorter than available rainfall data,
earlier land clearings in the QLD region cannot be identified spatially,
hence they are excluded from the analysis.

@@

Rainfall data for Australia (Jones, Wang, and Fawcett 2009) was obtained
from the Bureau of Meteorology. The data has been projected onto a
national 0.05\textdegree × 0.05\textdegree grid (approximately 5 km × 5
km). This gridded data set was generated from station observations using
an optimised Barnes successive correction technique. The Barnes
technique combines a weighted averaging process and defined
topographical information to estimate rainfall values between spatial
points (BoM 2009). The resulting data set provides additional
information for data-sparse areas like central Australia but reduces
information in the data-rich areas, such as southeast Australia where
station density is up to 20 per 100 \(\text{km}^2\). The data is
available on a monthly basis from 1900 to current. Here a subset of 36
years (1979 - 2015) was used. The study was conducted on monthly data,
as a land cover change effect on annual rainfall might be negligible but
can often found to be significant in particular months or seasons (e.g.
Otterman et al. 1990; Gaertner et al. 2001; Semazzi and Song 2001;
Oleson et al. 2004; Deo et al. 2009).

Large scale climate drivers are represented by various climatic indices.
The Southern Oscillation Index (SOI) is generally regarded as a good
predictor of Australian rainfall (Risbey et al. 2009; Chowdhury and
Beecham 2010; Westra and Sharma 2010), but its skill is weaker in some
parts of Australia. For example the Southern Annular Mode (SAM) is found
to be more important than ENSO in south Western Australia (Meneghini,
Simmonds, and Smith 2007). The suitability of each index for the regions
of interest was tested in section @ref(sec:reg\_model). The following
climate indices were used as candidate predictors for local rainfall.

@@

The step changes on rainfall were analysed using two different
statistical methods. A step change is not obvious in the time series
data of the rainfall residuals (Figure \ref{fig:ts-mean}), even though
the data is deseasonalised and detrended. Both methods in this paper
make use of a regression model to remove variability in rainfall due to
climatic factors to strengthen the tree cover change signal. In the
first method, the tree cover change was implemented as a factor variable
in the regression model. In the second method, a rank sum test (step
trend test), was applied to the model residuals after effects of other
major factors were removed. This assumes that after removal of all
climate variation, the vegetation cover change is the only factor
explaining the non-random pattern in the rainfall residuals.

@@

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific SSTs), IOD, SAM, MJO (Madden-Julian
oscillation) and blocking, in relation to Australian rainfall
variability. The MJO is a large scale eastward-propagating wave-like
disturbance in equatorial latitudes (Risbey et al. 2009). They
identified SOI as the most important index among all indices tested for
broad parts of Australia (including QLD and NSW/VIC) in almost any
season. In this study, four indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) were used as the
explanatory variables in the model for each study region.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation rather than the
correlation between trends in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall (as
shown in Figure \ref{fig:cor-rain-qld} and Figure
\ref{fig:cor-rain-nsw}), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the highest across all indices, outweighing the other ENSO indicators. IOD, had a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the highest correlation with rainfall, followed by the IOD. Both occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
\end{itemize}

The above findings are consistent with previous studies. Although some
indices are serially correlated with rainfall up to several months, the
lag zero events have the most significant correlation coefficients.
Concurrent climatic index series were generally found most useful in
rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
Metcalfe 2011). The correlations between the climatic indices and
rainfall for each individual season have also been tested, and the
results were similar.

>>>>

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\providecommand{\tightlist}{
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{lipsum}
\usepackage[unicode=true]{hyperref}
\usepackage{longtable}

\usepackage{lipsum} \usepackage{textcomp}

\title{Detecting the impact of land cover change on observed rainfall.}

\author[1]{Chun X. Liang}

\author[1]{Floris F. van Ogtrop}

\author[1]{R. Willem Vervoort}

\corrauthor[1]{R. Willem Vervoort}{\href{mailto:willem.vervoort@sydney.edu.au}{\nolinkurl{willem.vervoort@sydney.edu.au}}}

\affil[1]{Sydney Institute of Agriculture, The University of Sydney, NSW 2006}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

%

\begin{abstract}
Analysis of observational data to identify relationships between
rainfall and land cover change are difficult due to multiple
environmental factors that cannot be strictly controlled. In this study
we present a methodology to investigate the relationship using
statistical methods on data from best available sources at two sites in
Australia. Gridded data of rainfall and tree cover were used as
spatially corresponding local conditions. Large scale effects were
represented by climatic indicators, such as SOI and IOD. Regression
analysis and step trend tests were used to assess the effect of abrupt
land surface intervention. At a Queensland site, significant tree cover
change between 2002 - 2005 did not result in strong statistically
significant precipitation changes. On the other hand, results from a
bushfire affected NSW/VIC region suggests significant changes in the
rainfall. This indicates the method works better when a abrupt change in
the data can be clearly identified. The results from the step trend test
implied a positive relationship between the tree cover and the rainfall
at 0.1 significance level in both locations in data up to 2009. However,
high rainfall variability and possible regrowth meant that no signifcant
changes were observed in alonger time sereies to 2015.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}\label{introduction}
\addcontentsline{toc}{section}{Introduction}

Land use and land cover changes can lead to changes in the local
climate. Empirical and modelling studies have found that cloud types and
rainfall are correlated to large scale vegetation cover changes, such as
deforestation in the Amazon and in the Sahel (Chagnon and Bras 2005;
Pinto et al. 2009; Wang et al. 2009; Mei and Wang 2010; Kucharski, Zeng,
and Kalnay 2013; Pitman and Lorenz 2016) and afforestation in south
Israel (Otterman et al. 1990; Ben-Gai et al. 1998). Using airborne
measurement in Western Australia, Junkermann et al. (2009) showed a
significantly higher level of aerosols over an agricultural area
compared to an adjacent natural vegetation. They suggested that a
modification of aerosol concentrations due to deforestation could have
contributed to a reduction of local rainfall, as more, but smaller rain
droplets were observed. Nair et al. (2011) reported from the Bunny Fence
Experiment in Western Australia that local land use change altered the
synoptic west coast trough dynamics and surface roughness, and this
resulted in an observed rainfall decrease. Maximum temperatures were
also found to be sensitive to land cover change in eastern Australia
(McAlpine et al. 2007).

Overall the number of empirical studies analyzing changes to rainfall
due to land cover change from observational data is limited. Most of the
studies mentioned previously were either model simulations, or
comparisons of modelled data with observations. This is because there
are some fundamental experimental difficulties in both space (where does
evaporated water reappear as rainfall?) and in time (how much time does
it take for land cover change effects to appear or disappear?). In
addition, in many areas across the globe, rainfall variability is
related to a complex set of interactions, of which land use change might
only be a minor component.

Locally, there are two main sources that generate rainfall: moisture
from advective atmospheric transport; and local evapotranspiration
(Eltahir and Bras 1996; Bosilovich and Chern 2006; Dirmeyer, Brubaker,
and DelSole 2009; Gimeno et al. 2010). The local evapotranspiration
component is the component considered to be affected by land use change
(Eltahir and Bras 1996). According to Trenberth (1999), the contribution
of advective moisture partially depends on the availability of external
moisture and atmospheric transport. On the longer time scale, such as
monthly and annually, large scale atmospheric dynamics are affected by
large scale climate drivers. For example, many studies have reported
significant relationships between rainfall in large parts of Australia
and the El Niño-Southern Oscillation (ENSO) (Verdon et al. 2004; Risbey
et al. 2009; Speer, Leslie, and Fierro 2011). In contrast, local ET is
determined by local land surface characteristics, which influence local
scale atmospheric dynamics and hence the amount of rainfall, including
contribution from both main sources.

Although climate drivers demonstrate some capability to predict
Australian rainfall, there is still a large amount of unexplained
variance. Westra and Sharma (2010) pointed out that models based on
global sea surface temperature anomalies can only predict up to 14.7\%
of annual precipitation variance. Some of the remaining variance could
be due to land surface processes as suggested in studies predicting
local rainfall (e.g. Ma et al. 2011; Zeng et al. 2012; Pitman and Lorenz
2016; Saha, Dirmeyer, and Chase 2016). However, most are based on
modelling experiments and few empirical observational studies have been
reported. However, Pitman et al. (2004) found a good match between
observations and simulated rainfall changes in southwest Western
Australia, forced by land cover change. Timbal and Arblaster (2006) were
able to reproduce the rainfall decline in south west Australia by
including land cover influence. In addition, local land use change might
not be a primary, but is likely to be a secondary cause of rainfall
change (Nicholls 2006).

Therefore, the aim of this study is to use a statistical approach on
rainfall data at regional scales to investigate the cause and effect
relationship between land cover change and local rainfall, which is
demonstrated in many modelling studies. More specifically, we
hypothesize that a step change on the land cover on the surface will
cause a step change in the rainfall. To demonstrate this we study
changes in observed rainfall over time at a Queensland and NSW/Victoria
location where there are possible step changes in land cover change due
to land clearing and bush fires. The methodology uses statistical
approaches to identify changes in rainfall, which are subsequently
associated with land cover change through spatial comparison.

@@

In Australia, significant tree cover change has mainly occurred in the
north east and south east of the continent, as well as in the southwest
of Western Australia. According to the National Dynamic Land Cover
Dataset (DLCD) (Lymburner et al. 2010), most of these areas have
experienced decreases in the Enhanced Vegetation Index (EVI) post 2000
aws derived from satellite data. As an index for vegetation greenness,
the decreasing values indicate lower biomass over time in the tree cover
regions. The possible EVI reduction might be due to land clearing, bush
fires or drought.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/map_selreg} \caption{Selected study regions are highlighted by red rectangles in the main map (the red rectangle in the insert indicates the location of the main map). The types of tree cover in 2008 from the DLCD product is shown at the background. In site 1 (the QLD region), the tree cover is mostly sparse. In site 2 (the NSW/VIC region), many areas have open or close forest where the tree cover is denser.}\label{fig:selreg}
\end{figure}

Two regions were selected where significant tree cover change since 2000
was reported. The first region is located in south central Queensland
partly covering the north of the Murray Darling Basin (MDB) (site 1 in
Figure @ref(fig:selreg)). High rates of land clearing have been reported
in this region during the early 2000s (Department of Natural Resources
and Water 2007). The second study region is located at the border of New
South Wales and Victoria, and includes the Snowy Mountain ranges (site 2
in Figure @ref(fig:selreg)). Severe bush fires occurred in this area in
early 2003 (see Figure @ref(fig:figure2bushfire)). The 2003 bush fires
were the largest and the worst in the last 60 years (The State
Government of Victoria 2011). Two thirds of Kosciuszko national park was
heavily burned and regrowth was reported to be slow due to drought and
cold conditions (ABC News 2003), and the type of species in this region.
However, in the longer term, after an early high transpiration period a
recovery of pre-fire evapotranspiration would be expected (Kuczera
1987). For the purpose of this study, significant tree cover loss has
happened in both study areas in the last decade, either permanently or
temporarily.

\begin{figure}
\includegraphics[width=0.9\linewidth]{figures/bushfire_nswvic} \caption{Location of bushfires occurring in January 2003, in and around the NSW/VIC study region, as shown by the red pixels. The map shows the large area in the Kosciuszko national park that has been burned. Some locations in the southwest of teh Australian Capital Territory (ACT) have also experienced intensive bushfires.}\label{fig:figure2bushfire}
\end{figure}

The two regions have different characteristics. The QLD region is
partially grassland and subtropical, while the NSW/VIC region is mainly
within the temperate zone, under the Köppen classification. According to
Australian Bureau of Meteorology (BoM), the NSW/VIC region receives 1000
- 2000 mm rainfall annually, which is more than double the annual
rainfall in the QLD region. Evapotranspiration is similar in both
regions. Marine moisture and orographic effects are likely to be the
main contributors to rainfall in the southeast mountain areas of the
NSW/VIC region.

The land use and land cover characteristics in the two regions are also
different. In the Queensland region, the tree cover is sparse over most
of the area. The MODIS satellite tree cover data (discussed in more
detail in section 3) shows that tree cover in this region is generally
below 20\% of total ground area. Grazing is the main activity in this
region, with over 90\% of land used by the grazing industry (ABARES
2010). Our starting assumption is that the main cause of the EVI decline
over large part of the region is due to land clearing. Tree cover has
been cleared at a massive scale over the last decade, especially during
2002 - 2004. The reports from the Queensland Statewide Land Cover and
Trees Study (SLATS) (e.g. Department of Natural Resources and Mines
2005; Department of Science, Information Technology and Innovation 2017)
were used to investigate the time and location of the land clearing in
the QLD region.

The Kosciuszko national park is within the NSW/VIC region. Here tree
cover is denser with open or even closed forest (the tree cover
distribution is bimodal at 10 - 20\% and 60 - 70\%). The dominant
species in the alpine area are Snow Gum and large stand species such as
Alpine Ash and Mountain Gum in the sub-alpine area. These trees can
reach a great height but they take long time to grow. For example,
Alpine Ash would need about 20 years to mature. Although land clearing
is not a major issue in this region, it is vulnerable to fires and
drought. The MODIS burned area product, MCD45A1 (Roy, Lewis, and Justice
2002; Roy et al. 2005; Roy et al. 2008), was used to locate bush fires
areas in the NSW/VIC region, with a grid resolution of 500 m. MCD45A1
provides monthly burning information on all pixels, which helps to
pinpoint an abrupt event.

Due to the difference in nature of the land cover change in the two
regions, the post-change vegetation status is hypothesized to be
different as well (see Figure @ref(fig:figure3-tc-simple)).

@@

The overall hypothesis is that the effect of 2003 - 2004 land clearings
in the QLD region and the 2003 bush fires in the NSW/VIC region cause a
step change in the local rainfall. The actual tree cover change during
this time at the pixel level was derived from the 15-year MODIS data
(discussed below). The difference of tree cover before and after the
land disturbance was tested using a Student's t-test. As the length of
the tree cover data is shorter than available rainfall data, earlier
land clearing in the QLD region cannot be identified spatially, hence
they are excluded from the analysis.

@@

The SILO Rainfall product data for Australia was used (Jeffrey et al.
2001) (available online at \url{https://silo.longpaddock.qld.gov.au/}))
. The data has been projected onto a national 0.05\textdegree by
0.05\textdegree grid (approximately 5 km by 5 km). This gridded data set
was generated from station observations using spline interpolation and
kriging (Jeffrey et al. 2001). The data has been compared to other
gridded products and observed data and is generally of high quality
(Tozer, Kiem, and Verdon-Kidd 2009; Tozer, Kiem, and Verdon-Kidd 2012).
The data is available on a daily and monthly basis from 1889 to current.
Here a subset of 36 years (1979 - 2015) was used. The study was
conducted on monthly data, as a land cover change effect on annual
rainfall might be negligible, but can often be significant in particular
months or seasons (e.g. Otterman et al. 1990; Gaertner et al. 2001;
Semazzi and Song 2001; Oleson et al. 2004; Deo et al. 2009).

Large scale climate drivers are represented by various climatic indices.
The Southern Oscillation Index (SOI) is generally regarded as a good
predictor of Australian rainfall (Risbey et al. 2009; Chowdhury and
Beecham 2010; Westra and Sharma 2010), but its skill is weaker in some
parts of Australia. For example the Southern Annular Mode (SAM) is found
to be more important than ENSO in south Western Australia (Meneghini,
Simmonds, and Smith 2007). The testing of the suitability of each index
for the regions of interest is described in a later section. The
following climate indices were used as candidate predictors for local
rainfall.

@@

As an initial analysis a simple boxplot and t-test is used to analyse
whether there is a significant change in tree cover in time, and before
and after the suspected change in the regions.

To assess the actual causal relationship between the tree cover and the
rainfall a flexible regression model is applied (discussed in detail
below). A step change is not directly obvious in the time series of the
rainfall anomalies (Figure @ref(fig:ts-mean)) for both regions, even
though the data is deseasonalised and detrended. In this study we apply
different statistical methods to analyse the effect of tree cover change
on rainfall. Both methods make use of a regression model to remove
year-on-year variability in rainfall to strengthen the tree cover change
signal.

In the first method, the tree cover change is implemented as a factor
variable in the regression model, and the significance of this variable
is tested. In the second method, a rank sum test (step trend test), is
applied to the regression model residuals after effects of other major
factors were removed. This assumes that after removal of all climate,
long term linear trend and seasonal variation, the vegetation cover
change is the only factor explaining the non-random pattern in the
rainfall residuals.

@@

As highlighted in the introduction, the Australian climate is influenced
by sea surface temperatures in the tropical Pacific and Indian Oceans,
as well as pressure systems in the Southern Ocean (BoM 2012b). Risbey et
al. (2009) compared five large-scale drivers, including ENSO (measured
by SOI and the Tropical Pacific Sea surface temperatures (SSTs)), IOD,
SAM, MJO (Madden-Julian oscillation) and blocking, in relation to
Australian rainfall variability. The MJO is a large scale
eastward-propagating wave-like disturbance located around equatorial
latitudes (Risbey et al. 2009). They identified SOI as the most
important index among all climate indices tested for broad parts of
Australia (including QLD and NSW/VIC) in almost any season. In this
study, four climate indices were selected from the main climatic
indicators (see section \protect\hyperlink{Data}{Data}) and used as the
explanatory variables in the model for each study region. A further
complicating factor is the influence of the ``millenium drought'' over
the study period and in particular the change to wet conditions in 2010
- 2011 (Dijk and Viney 2013). Therefore, the spatially averaged monthly
rainfall in the Murray Darling Basin (MDB, downloaded from
\href{http://www.bom.gov.au/web01/ncc/www/cli_chg/timeseries/rain/allmonths/mdb/latest.txt}{the
Bureau of Meteorology}) was used to explain the year-on-year variation
in the rainfall in the regions. As climate alone are unlikely able to
fully explain the temporal variation in rainfall in the regions,
including the drought (Dijk and Viney 2013; Westra and Sharma 2010),
this variable was included. Since both regions at least partly overlap
with the MDB, the average rainfall across was assumed to be a useful
explaining variable.

Correlations between rainfall and each climate index were analysed.
Rainfall in each study region was first deseasonalised and detrended
using the seasonal decomposition function \texttt{ds} in the package
\texttt{deseasonalise} in R (R Core Team, 2018). Using detrended data
gives a better indication of the underlying correlation compared to the
correlation between trends in the data (Smith and Timbal 2012). The
cross-correlations between the deseasonalised and detrended rainfall and
the climatic indices were tested using the Pearson's product moment
correlation method, assuming the relationships are linear. Although the
optimal technique for exploring the correlation with each index could be
different as described in Risbey et al. (2009), the Pearson's method was
applied to all indices for consistency. Because the PDO describes the
multi-decadal SST with lower frequency (MacDonald and Case 2005;
Zanchettin et al. 2008; Kamruzzaman, Beecham, and Metcalfe 2011),
instead of 37-year rainfall data, a longer period (115 years, from 1900
to 2015) was used to estimate the correlation with PDO, up to lag 24.
For the other indices, the 37-year data was used.

@@

Based on the correlation between the climatic indices and rainfall in
the regions (as shown in Figure @ref(fig:cor-rain-qld) and Figure
@ref(fig:cor-rain-nsw)), it can be concluded that:

\begin{itemize}
  \setlength{\itemsep}{0cm}
  \setlength{\parskip}{0cm}
  \item In QLD, the correlation between rainfall and SOI at zero time lags is the strongest across all indices, outweighing the other ENSO indicators. IOD and PDO have a weak influence in QLD.
  \item In NSW/VIC, again the SOI has the strongest correlation with rainfall, followed by the IOD. For both, the strongest correlations occur at the zero time lags. 
  \item in both cases PDO had the weakest correlations, and this factor was not further considered as a predictor.
  \item in general, for the better correlated indices, strongest correlations occured at zero time lags.
\end{itemize}

The above findings are consistent with previous studies. Although some
indices are serially correlated with rainfall up to several months, the
lag zero events have the greatest correlation coefficients. Using
multiple climatic index series was generally found most useful in
rainfall prediction (e.g. Risbey et al. 2009; Kamruzzaman, Beecham, and
Metcalfe 2011).
